Diversity and evolution of centromere repeats in the maize genome

### 2014-05-04 centc.tex
done with discussion.

\jri{check tense throughout. i think some is past and some is present} 
Our analysis of centromere repeat diversity across the maize genome identifies thousands of copies exhibiting tremendous diversity. 
Nonetheless, while we can cluster the repeats into groups of related sequences, these groups seem to have little relation to current or ancient maize chromosomes (Figure \ref{pcoa}).  
We see no evidence of chromosome specific repeats such as that seen in X and Y \citep{Kawabe2005, Pontes2004}, suggesting the presence of mechanisms homogenizing repeats across centromeres on different chromosomes. We further show that Cent4, while specific to chromosome 4, appears to be a poorly characterized tandem repeat or retroelement, but is not associated with the centromere.

Instead of chromosome-specific groups, we find that virtually all the large arrays of CentC in the maize  genome derive from the maize1 subgenome, which is known to have lower gene loss and higher average expression than the maize subgenome 2 \cite{Schnable2011}. \jri{is it maize1 or maize subgenome 1? you use both here} 
One possible explanation for this clear bias is that differential expression between subgenomes affects centromere function.  
Regulatory microRNA’s corresponding to centromeric repeats have been observed\citep{ReinhartBartel2002}, and perhaps the silenced centromere became more susceptible to deletion just like other parts of subgenome2.  
We see no evidence of a more rapid decay of CentC sequences in subgenome2, but we likely have poor ability to detect deletion of entire tandem arrays.  
An alternative explanation is may simply be that the subgenome1 parent had more or larger CentC arrays, leading to an unequal representation in the ancient polyploid ancestor of maize.

@@

%!RELATEDNESS
Our sequence comparison of CentCs also enabled us to explore the relationship between genetic and physical distance between repeats.  
Using the well-assembled centromeres on chromosomes 2 and 5, we found spatial autocorrelation among repeats, and we observe genome-wide that many CentC’s within an array fall into the same genetic cluster.  
These observations are consistent with the simple idea that most repeats arise due to tandem duplication, and long distance transposition of CentC, while necessary to homogenize repeats across chromosomes \citep{Shi2010}, is relatively uncommon.

One unusual result from our sequence comparison was the finding that pairs of CentC on different chromosomes may be identical or nearly so.  Our simulations suggest that this is plausible due to homoplasmy alone under realistic assumptions about mutation rate and divergence time.  We speculate that the vast majority of the CentC’s exist as a result of old duplications and that mutations have had a long time to accumulate. Roughly 80\% of the CentC repeats have their closest genetic relative on the same chromosome, as expected under a model of tandem duplication.  However, only 14\% of closest genetic pairs are found within 10KB of each other, suggesting that their tandem duplication is old and that most CentC’s have persisted within the genome for a long time.

>>>>

\jri{check tense throughout. i think some is past and some is present. also fill in species names X and Y and add CITEs} 
Our analysis of centromere repeat diversity across the maize genome identifies thousands of copies exhibiting tremendous diversity. 
But while we can cluster the repeats into groups of related sequences, these groups seem to have little relation to current or ancient maize chromosomes (Figure \ref{pcoa}).  
We see no evidence of chromosome specific repeats such as that seen in X and Y \citep{Kawabe2005, Pontes2004}, suggesting the presence of a mechanisms that homogenizes repeats across centromeres on different chromosomes. 
We further show that Cent4, once thought to be a chromosome-specific centromere repeat (CITE), appears to be a poorly characterized tandem repeat or nonautonomous retroelement, but is not associated with the centromere.  
Inter-chromosomal transposition was postulated by \citet{Shi2010} as an explanation for marker genotypes showing evidence of recombination, and such transposition could also explain the similarity of repeats across all chromosomes.

Instead of chromosome-specific groups, we find that virtually all the large arrays of CentC in the maize  genome derive from one of the two ancestral genomes present in modern day \emph{Zea} (Figure \ref{circos}).  
One possible explanation for this result is that centromere inactivation mirrors the differential expression seen between the subgenomes.  
Centromere inactivation could then lead to differential loss of CentC clusters, such that large clusters remain only in the active centromeres.
Large CentC clusters are clearly biased in favor of subgenome 1, which is known to have lower gene loss and higher average expression than maize subgenome 2 \cite{Schnable2011}. 
Consistent with this idea, \citet{ReinhartBartel2002} find evidence of regulatory microRNA’s corresponding to centromeric repeats.   
Of course, an alternative explanation is simply that the subgenome 1 parent had more or larger CentC arrays, leading to an initial unequal representation in the ancient polyploid ancestor of maize.

@@

%!RELATEDNESS
Our sequence comparison of CentCs also enabled us to explore the relationship between genetic and physical distance among repeats.  
Using the well-assembled centromeres on chromosomes 2 and 5, we found spatial autocorrelation of relatedness among repeats, and we observe genome-wide that many CentC’s within an array fall into the same genetic cluster.  
These observations are consistent with the simple idea that most repeats arise due to tandem duplication, and long distance transposition of CentC, while necessary to homogenize repeats across chromosomes \citep{Shi2010}, is relatively uncommon.

One unusual result from our sequence comparison was the finding that pairs of CentC on different chromosomes share very high sequence similarity.  
Our simulations suggest that, under realistic assumptions about mutation rate and divergence time, such a pattern is possible due to homoplasious mutation alone.  
Roughly 80\% of the CentC repeats have their closest genetic relative on the same chromosome, as expected under a model of tandem duplication, but only 14\% of closest genetic pairs are found within 10KB of each other.
We speculate that the vast majority of  CentC’s in the genome are thus a result of relatively old tandem duplications, and that sufficient time has occurred since duplication for rearrangements and mutations to break up patterns of identical tandem repeats.
